---
title: "FA Homework 3"
author: "John Navarro"
date: "July 8, 2017"
output: pdf_document
---

This assignment helps understanding linear models for time series
```{r, warning=FALSE, include=FALSE}
library(tseries)
library(forecast)
library(TSA)
```


## 1. Exercise 2 on page 125

Use the file m-dec125910-6111.txt
```{r}
# read in data
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/FinancialAnalytics/Week3_TimeSeries"
da=read.table(paste(datapath,"m-dec125910-6111.txt",sep="/"),header=T)
head(da)
dim(da)
```


### For CRSP portfolios of Decile 2 and Decile 10 returns test null hypothesis that the first 12 lags of autocorrelations equal zero with 5% level

```{r}
# select the Decile 2 returns and transfer to log returns
d2 <- da$dec2
d2.ts <- ts(d2,frequency = 12)
lnd2 <- log(d2.ts+1)
# Plot the time series and its ACF for simple returns
par(mfcol=c(2,1))
plot(d2.ts, xlab="year", ylab="returns",type="l")
title(main = "Simple returns of Decile 2")
acf(d2.ts, lag=24)
# Plot the time series and its ACF for log returns
par(mfcol=c(2,1))
plot(lnd2, xlab="year", ylab="log returns",type="l")
title(main = "Log returns of Decile 2")
acf(lnd2, lag=24)

```

The simple returns and log returns of Decile 2 look similar and so do their ACF plots, we will use a Portmanteau Test to check for autocorrelations in the first 12 lags.

```{r}
# Test the H0 that the first 12 lags of autocorrelations of decile 2 equal zero
# Use the Ljung Box test for the simple returns
Box.test(d2.ts, lag=12, type="Ljung")
# Use the Ljung Box test for the log returns
Box.test(lnd2, lag=12, type="Ljung")

```
In the case of simple returns for Decile 2, we cannot reject the null hypothesis of no serial correlations in the returns at the 5% level. Although, the p-value is smaller in the log returns case, we still cannot reject the null hypothesis of no serial correlations in the log returns at the 5% level.

```{r}
# select the Decile 10 returns and transfer to log returns
d10 <- (da$dec10)
d10.ts <- ts(d10, frequency = 12)
lnd10 <- log(d10.ts+1)
# Plot the time series and its ACF for simple returns
par(mfcol=c(2,1))
plot(d10.ts, xlab="year", ylab="returns",type="l")
title(main = "Simple returns of Decile 10")
acf(d10.ts, lag=24)
# Plot the time series and its ACF for log returns
par(mfcol=c(2,1))
plot(lnd10, xlab="year", ylab="log returns",type="l")
title(main = "Log returns of Decile 10")
acf(lnd10, lag=24)
```

Again the simple returns and log returns of Decile 10 look similar. Again, we use a Portmanteau Test to check the first 12 lags for autocorrelations of 0.

```{r}
# Test the H0 that the first 12 lags of autocorrelations of decile 10 equal zero
# Use the Ljung Box test for the simple returns
Box.test(d10.ts, lag=12, type="Ljung")
# Use the Ljung Box test for the log returns
Box.test(lnd10, lag=12, type="Ljung")
```

In this case we can see that both results give us p-values that are significant at the 5% level. Therefore, we can reject the null hypothesis and say that there exists serial correlations in the monthly simple returns of the Decile 10 data. Similarly, we can draw the same conclusion for the log returns of the data as well.

### Fit ARMA model for returns of Decile 2, perform model checking and write down the fitted model

```{r}
d2ts <- ts(d2, frequency = 12)
par(mfcol=c(1,2))
acf(d2ts, lag =35)
pacf(d2ts, lag =35)
eacf(d2ts) # suggests the best fit is an MA(1)
auto.arima(d2ts) # suggests the best fit is an MA(1)
(m11 <- arima(d2ts, order=c(1,0,1)))
(m1 <- arima(d2ts, order=c(0,0,1)))
```
Here we can see that the MA(1,1) model gives us an AIC of-1987.68, while the MA(1) model improves the AIC slightly to -1988.08

Here are the equations for both models:
ARMA (1,1)
(1+0.4039B)xt = 0.0093 +(1-0.5265B)at
MA(1)
Xt=0.0093 +(1-0.1307)at

Model checking using Ljung-Box and diagnostics
```{r}
Box.test(m11$residuals, lag=12, type="Ljung")
Box.test(m1$residuals, lag=12, type="Ljung")
tsdiag(m11)
tsdiag(m1)

```
 Here we can see that the p-values are not significant, and we cannot reject the null hypothesis that the residual have autocorrelation.

### Use the fitted model to produce 1- to 12-step ahead forecasts of the series and the associated standard errors of forecasts.

```{r}
# Using the ARMA(1,1) model
prd <- predict(m11,12)
cbind(Predicted=as.vector(prd$pred),Standard.Error=as.vector(prd$se))
# Using the MA(1) model
prd <- predict(m1,12)
cbind(Predicted=as.vector(prd$pred),Standard.Error=as.vector(prd$se))
```

Since we don't have any out of sample data (post September 2011) for Decile 2, we cannot determine the accuracy of these predictions. However, by looking at them, we see that after the first forecast, the MA(1)  model returns the unconditional mean value as a prediction. While in the ARMA(1,1) forecasts, we can see the half life nature of the correlations. The forecasts will also converge to the unconditional mean value.

## 2. Exercise 4 on page 126

Consider the monthly yields of Moody's Aaa & Baa seasoned bonds from
January 1919 to November, 2011. The data are obtained from FRED of Federal
Reserve Bank of St. Louis. Consider the log series of monthly Aaa bond yields.
Build a time series model for the series, including model checking.

```{r}
# read in data
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/FinancialAnalytics/Week3_TimeSeries"
da=read.table(paste(datapath,"m-aaa-1911.txt",sep="/"),header=T)
tail(da)
dim(da) #1115 4
lnyld <- log(da$yield/12/100)
plot(lnyld)
# Build a time series
yld.ts <- ts(lnyld, frequency=12)
par(mfrow=c(1,2))
acf(yld.ts, lag =35)
pacf(yld.ts, lag =35)
```

The plot of the time series is not stationary. So we may have to difference it.
ACF looks like a non-stationary series

```{r}
eacf(yld.ts) # suggests the best fit is an ARMA(1,1)
auto.arima(yld.ts) # suggests the best fit is an ARIMA(0,1,1)
# We chose to build an ARIMA(0,1,1) model
m.011 <- arima(lnyld, order=c(0,1,1))
summary(m.011)
# Check the residuals
Box.test(m.011$residuals, type="Ljung")
tsdiag(m.011)
```
The pvalue is not significant, so we cannot reject the null hypothesis of no serial correlations.



