---
title: "Trading Club Random Forest"
author: "John Navarro"
date: "November 9, 2017"
output: pdf_document
---
References

TTR package
https://cran.r-project.org/web/packages/TTR/TTR.pdf

Khaidem - Predicting the direction of stock market prices using random forest
https://arxiv.org/pdf/1605.00003.pdf

The goal of the project is to adapt the methodology from the reference paper and use random forests to predict the direction of price movement for Bitcoin.

This project can be expanded to include multiple cryptocurrencies, with the goal to identify the most similar pairs in regards to direction of price movement.

```{r setup, include=FALSE}
library(quantmod)
library(TTR)
```

Import data
```{r}
dataPath="C:/Users/JohntheGreat/Documents/MSCA/TradingClub"
data = read.csv( file=paste(dataPath,"BTC.csv",sep="/"))

```

```{r}

btc.data <- data[,c(2,4)]
diff<- diff(btc.data$last)
btc.data <- btc.data[-1,]
btc.data <- cbind(btc.data,diff)
head(btc.data)
dim(btc.data)
```

Using Package TTR
```{r}
btc.rsi <- RSI(btc.data$last, 14)
head(btc.rsi, 20)
# need to remove the first 14 elements of rsi (NA values)
btc.data <- cbind(btc.data, btc.rsi)
```
macd
```{r}
btc.macd <- MACD(btc.data$last,12,26,9)
#head(btc.macd,35)
# need to remove first 33 elements of NAs
btc.data <- cbind(btc.data, btc.macd)
btc.data[30:65,]
```

