---
title: "FinancialAnalysisHomework1Navarro"
author: "John Navarro"
date: "June 23, 2017"
output: pdf_document
---

### 1. Read Chapter 1 of the book
### 2. Download and analyze excess returns of S&P 500

Calculate continuous daily excess returns of SP500("^GSPC") for the period from 1/1/2014 until 12/31/2014 using overnight Fed Funds rates as risk-free rates.
```{r, warning=FALSE, include=FALSE}
# Install packages
suppressWarnings(library(quantmod))
library(moments)
```

```{r}
# Read in the data
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/FinancialAnalytics/Week1_Introduction"
Curr <- read.csv(file=paste(datapath, 'GBP_JPY.csv', sep='/'))
SP500 <- read.csv(file = paste(datapath, 'FSP500.csv', sep='/'))
# Check the data
head(SP500)
# Calculate the simple daily return
SP500$simple.excess.return <- SP500$RIFSPFF_N.B - SP500$SP500Returns
#Calculate the continuous excess return
SP500.cont.excess.return <- diff(log(SP500$simple.excess.return))
# Plot the continuous excess return
plot(SP500.cont.excess.return, type="l", ylab="Excess returns of SP500", xlab = "Date", xaxt="n", main="Continuous Excess Daily Returns of S&P 500")
axis(1, at=10*(1:25), labels=SP500[10*(1:25),1])

```

### 3. Download and analyze exchange rates

Answer the same questions as in Exercise 5 on page 37 as a refresher of statistical analysis skills. Try to do it without using R demo code from the book

1. Download GBP/USD and USD/JPY exchange rates from Oanda using quantmode
```{r}
#getSymbols("GBP/USD",src="oanda", from= "2017-01-02", to= Sys.Date())
#getFX("USD/JPY")
```
Using the above code gives the following error. "Error in charToDate(x) : character string is not in a standard unambiguous format"
So I gathered closing FX data from Bloomberg for both GBP/USD and USD/JPY for 1/2/2007 to 11/30/2011. These dates were from Exercise 5 in the textbook.

2. Calculate daily log returns of both exchange rates
```{r}
head(Curr)
GBP.daily.log.return <- diff(log(Curr$GBPClose))
head(GBP.daily.log.return)
JPY.daily.log.return <- diff(log(Curr$JPYClose))
head(JPY.daily.log.return)

```

3. Calculate min, mean, sd, skewness, Kurtosis, and max of log returns for both exchange rates

```{r}
c(GBP.min = min(GBP.daily.log.return),
  GBP.mean = mean(GBP.daily.log.return),
  GBP.sd = sd(GBP.daily.log.return),
  GBP.skew = skewness(GBP.daily.log.return),
  GBP.kurt = kurtosis(GBP.daily.log.return),
  GBP.max = max(GBP.daily.log.return))
```

```{r}
c(JPY.min = min(JPY.daily.log.return),
  JPY.mean = mean(JPY.daily.log.return),
  JPY.sd = sd(JPY.daily.log.return),
  JPY.skew = skewness(JPY.daily.log.return),
  JPY.kurt = kurtosis(JPY.daily.log.return),
  JPY.max = max(JPY.daily.log.return))
```

4. Test hypothesis H0: u=0 against alternative H0: u!=0
```{r}
t.test(GBP.daily.log.return)
t.test(JPY.daily.log.return)
```
 In both cases the p- values are not significant. So we cannot reject the null hypothesis in either case. We say that the means are not significantly different from zero.
